Open pNets

### 2020-10-16 2020-Newsemantics/WeakBisim.tex
cleaner semantics?

\documentclass{lmcs}
%\overfullrule=1mm
\usepackage[T1]{fontenc}
\usepackage{lmodern}
%\usepackage{geometry}                % See geometry.pdf to learn the layout options. There are lots.
%\geometry{a4paper}                   % ... or a4paper or a5paper or ...
%\geometry{landscape}                % Activate for for rotated page geometry
%\usepackage[parfill]{parskip}    % Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}
\usepackage{DotArrow}
\usepackage{hyperref}
\usepackage{listings}
\lstset{basicstyle=\small}

@@

\pagestyle{plain}

@@

\title{Compositional equivalences based on Open pNets
%\thanks{This work was partially funded by the Associated Team FM4CPS
%  between INRIA and ECNU, Shanghai}
}
\author{Rab\'ea Ameur-Boulifa}
\address{LTCI, T\'el\'ecom Paris, Institut Polytechnique de Paris, France
}         
\email{rabea.ameur-boulifa@telecom-paris.fr}

\author{Ludovic Henrio}
\address{Univ Lyon, EnsL, UCBL, CNRS, Inria,  LIP, F-69342, LYON Cedex 07, France.}
\email{ludovic.henrio@cnrs.fr}

\author{Eric Madelaine}
\address{INRIA Sophia Antipolis M\'edit\'erann\'ee, UCA, BP 93, 06902 Sophia Antipolis, France}
\email{eric.madelaine@inria.fr}
%\date{}                                           % Activate to display a given date or no date

		%\section{}
%\subsection{}

\begin{abstract}

Establishing equivalences between programs or system is crucial both for verifying correctness of programs, by establishing that two implementations are equivalent, and for justifying optimisations and program transformations, by establishing that a modified program is equivalent to the source one. There exist several equivalence relations for programs, and bisimulations are among the most versatile of these equivalences. Among bisimulation relations one distinguishes strong bisimulation, that requires that each action of a program is simulated by a single action of the equivalent program, a weak bisimulation that  is a coarser relations, allowing some of the actions to be
  invisible or internal moves, and thus not simulated by the equivalent program.

pNets is a generalisation of automata that includes parameters, and hierarchical composition. Open pNets are pNets with holes, i.e. placeholders inside the hierarchical structure  that can be filled by sub-systems. Reasoning on open pNets allows us to reason on open systems.
  Open pNets have a notion of \emph{synchronised actions}
  generalizing the usual internal actions (e.g. $\tau$ of CCS, or $i$
  in Lotos).

This article defines bisimulation relations  for the comparison of systems specified as pNets.
 We first define a strong bisimulation for open pNets.
  In practice, as happens in process algebras, strong bisimulation is too strong, and
  we need to define some coarser relations, taking into account
  invisible or internal moves.
  We then define an equivalence relation similar to
  the classical \emph{weak bisimulation}, and study its properties.
Among these properties we are interested in compositionality: If two systems are proven equivalent they will be undistinguishable by their context, and they will also be undistinguishable when their holes are filled with equivalent systems.
The article is illustrated with a transport protocol running example; it shows the characteristics of our formalism and our bisimulation relations.
\end{abstract}

@@

\section{Introduction}

In the nineties, several 
works extended the basic behavioural models based on labelled
transition systems to address value-passing or parameterised systems, using
various symbolic encodings of the
transitions~\cite{deSimone85,Larsen87}. 
In \cite{IngolfsdottirL:2001,HennessyLin:TCS95}, Lin, Ingolfsdottir and Hennessy developed a full hierarchy of
bisimulation equivalences, together with a proof system, for value passing CCS,
including notions of symbolic behavioural semantics and various symbolic 
bisimulations (early and late, strong and weak, and their congruent versions). They also extended this work to models with explicit assignments \cite{Linconcur96}.
Separately J. Rathke~\cite{HennessyRathke:TCS98} defined another symbolic semantics for 
a parameterised broadcast calculus, together with strong and weak bisimulation
equivalences, and developed a symbolic model-checker based on a tableau
method for these processes. 30 years later, no
practical verification approach and no verification platform are
using this kind of approaches to provide proof methods for
value-passing processes or open process expressions.

This article provides a theoretical background that allows us to implement such a verification 
platform. We build upon the concept of pNets that allowed us to give a behavioural semantics 
of distributed components and verify the correctness of distributed applications in the past 15 years. pNets  is a
low level semantic framework for expressing the behaviour of various
classes of distributed languages, and as a common internal format for
our tools.  pNets allow the
specification of parameterised hierarchical labelled
transition systems:  labelled transition systems with parameters can be
combined hierarchically.

We develop here a semantics for a model of interacting processes with parameters and holes. 
The main interest of our symbolic approach is to define a method to
prove properties directly on open structures; these properties will then be preserved
by any correct instantiation of the holes.
As a consequence, our model allows us to reason on composition
operators as well as on full-size distributed systems. The parametric
nature of the model and the properties of compositionability of the
equivalence relations are thus  the main
strengths of  our approach.

\subsubsection*{Previous Works and Contribution}
While most of our previous works relied on closed, fully-instantiated semantics~\cite{BBCHM:article2008,AmeurBoulifa2017,HKM-FASE16}, it is only recently that we could design a first version of a  parameterised semantics for pNets with a strong bisimulation equivalence~\cite{henrio:Forte2016}. This article builds upon this previous parameterised semantics and provides a cleaner version of the semantics with a slightly simplified formalism. It also adds a notion of global state to automata. The slight simplification of the formalism allowed us to adopt a much cleaner presentation where the theory for open automata. The algebra used to define the semantics of pNets with holes, can be studied independently from its application to pNets. Also, in \cite{henrio:Forte2016} the study of compositionability was only partial, and in particular the proof that bisimulation is an equivalence is one new contribution of this article and provides a particularly interesting insight on the semantic model we use.
The new formalism allowed us to extend the work and define weak bisimulation  for open automata, which is entirely new. This allows us to define a weak bisimulation equivalence for open pNets with valuable properties of compositionality. 
To summarise, the contribution of this paper are the following:
\begin{itemize}
\item The definition of open automata: an algebra of parameterised automata with holes, and a strong bisimulation equivalence. This is an adaptation of~\cite{henrio:Forte2016} with an additional property stating that strong bisimulation equivalence is indeed an equivalence relation.
\item A semantics for open pNets expressed as translation to open automata. This is an adaptation of~\cite{henrio:Forte2016} with a complete proof that strong bisimulation is compositional.
\item A theory of weak bisimulation for open automata, and its properties. It relies on the definition of weak open transitions that are derived from transitions of the open automaton by concatenating invisible action transitions with one (visible or not) action transition. The precise and sound definition of the concatenation is also a major contribution of this article.
\item A resulting weak bisimulation equivalence for open pNets and a simple static condition on synchronisation vectors inside pNets that is sufficient to ensure that weak bisimulation is compositional.
\item An illustrative example based on a simple transport protocol, showing the construction of the weak open transitions, and the proof of weak bisimulation.
\end{itemize}

\subsubsection*{Structure}
This article is organised as follows. Section~\ref{sec:notations}
provides the definition of pNets and introduces the notations used in
this paper, including the definition of open pNets. 
Section~\ref{sec:OT} defines open automata, i.e. automata
with parameters and transitions conditioned by the behaviour of
``holes''; a strong bisimulation equivalence for open automata is also
presented in this section. 
Section~\ref{section:op-semantics} gives
the semantics of open pNets expressed as open automata, and states
compositional properties on the strong bisimulation for open
pNets. 
Section~\ref{sec:weak} defines a weak bisimulation
equivalence on open automata and derives weak bisimilarity for pNets,
together with properties on compositionality of weak bisimulaiton for
open pNets. 
Finally, Section~\ref{sec:RW} discusses related works and Section~\ref{section:conclusion} concludes the
paper.

@@

For this property to be valid even if the substitution does not operate on all variables,
we define the composition operation as follows: 
%\TODO{version d eric reprise et corrigee}
%$(x_k\gets e_k)^{k\in K}\shortotimes \Post =  (x_k\gets e_k\subst{\Post})^{k\in K}$
%\ERIC{J'ai un probleme ici... quand $Post'$ agit sur des vars de $t$ qui ne sont pas substitutees par $Post$, alors les 2 defs de $\shortotimes$ ci-dessus ne sont pas equivalentes:}\\
%$F(x,z)(\subst{x\gets z}\shortotimes\subst{z\gets 1})=F(x,z)(\subst{x\gets 1})=F(1,z)$\\
%\ERIC{Alors que}\\
%$(F(x,z)\subst{x\gets z})\subst{z\gets 1})=F(z,z)\subst{z\gets 1}=F(1,1)$
%
%\ERIC{Une definition plus correcte serait peut-etre:}\\

\[(x_{k}\gets e_{k})^{k\in K}\shortotimes (x'_{k'}\gets e'_{k'})^{k'\in K'} =  
(x_{k}\gets e_{k}\subst{(x'_{k'}\gets e'_{k'})^{k'\in K'}})^{k\in K} \cup (x'_{k'}\gets e'_{k'})^{k'\in K''}\]
where $K''=\{k'\in K'|x'_{k'}\not\in\{x_k\}^{k\in K}\}$
%
%\LUDO{En fait comme on supose qu il y a pas de variable globale et donc que les post doivent affecter toutes les variables ce contre-exemple n est pas applicable! apres on peut definir comme ca si on veut ceinture et bretelles mais ca alourdit un peu la definitions}
%
%\ERIC{Heu Non... 
%  En pratique, dans la composition de posts de la regle WT3, les posts assignent bien toutes les variables de leur OT (strong), mais dans la composition, on fini par avoir des trous... Voir page 21-22. On est tombes tous les deux separemment, Rabea et moi, sur ce genre de cas.}
% 
%\LUDO{euuh ca doit etre lie qu fait qu il manque des ec leftarrowec dans les selfloop de l'OA de spec. Plutot que changer la def et les preuves je propose de rajouter la condition $vars(s')=dom(Post)$ pour tous les OAs ce qui ne change rien a part pour les selfloop que l on s autorisait apparemment a ecrire en omettant $x:=x$ ... au pire l'identite est un post toujours correct et valide}

\subsection{Parameterised Networks (pNets)}
\label{section:pnets}

@@

\subsection{Running Example}
To illustrate this work, we use a simple communication protocol, that provides safe transport of data between two processes, over unsafe media.

Figure \ref{SimpleProt:Spec} (left) shows the example principle, which corresponds to the hierarchical structure of a pNet: two unspecified
processes $P$ and $Q$ (holes) communicate messages, with a data value
argument, through the two protocol entities. Process $P$ sends an $\texttt{p-send(m)}$ message to the $\symb{Sender}$; this communication is denoted as $\texttt{\nounderline{in(m)}}$% , where underline actions represent internal actions that cannot be further synchronised with the environment, similarly to the $\tau$ of CCS. In our formalisation we will also denote these actions $\tau$ for simplicity of notations but the examples are easier to read when invisible actions have significant names}
.
At the other end, process $Q$ receives the message from the $\symb{Receiver}$. The holes $P$ and $Q$ can also have other interactions with their environment, represented here by actions $\texttt{p-a}$ and $\texttt{q-b}$. The underlying network is modelled by a medium entity transporting messages from the sender to the receiver, and that is able to detect transport errors and signal them to the sender. The return $\symb{ack}$ message from $\symb{Receiver}$ to $\symb{Sender}$ is supposed to be safe. The final transmission of the message to the recipient (the hole $Q$) includes the value of the ``error counter'' $ec$.

Figure \ref{SimpleProt:Spec} (right) shows a graphical view of the pNet that specifies the system:
the full system should be equivalent (e.g.  weakly bisimilar)
to the two processes connected simply through a perfect medium.  
The pNet has a tree-like structure.  The root node of the tree {\it SimpleSystemSpec} is the top level of the pNet structure. It acts as the parallel operator. It consists of three nodes: two holes $P$ and $Q$ and one sub-pNet, denoted {\it PrefectBuffer}. Nodes of the tree are synchronized using four synchronization vectors, that  express the possible synchronizations between the parameterised actions of a subset of the nodes. For instance, in the vector "$< \texttt{p-send(m)},\texttt{in(m)},\_> \rightarrow \texttt{\nounderline{in(m)}}$" only $P$ and {\it PrefectBuffer} nodes are involved in the synchronization. The synchronization between these processes occurs when  process $P$  performs $\texttt{p-send(m)}$  action sending a message, and  the {\it PrefectBuffer} accepts the message through an $\texttt{in(m)}$ action  at the same time; the result that will be returned at upper level is the action  $\texttt{\nounderline{in(m)}}$.

Figure \ref{SimpleProt:Impl} shows the pNet model of the protocol implementation. When the $\symb{Medium}$ detects an error (modelled by a local $\tau$ action), it sends back a $\texttt{m-error}$ message to the $\symb{Sender}$. The $\symb{Sender}$ increments its local error counter $ec$, and resends the message (including $ec$) to the $\symb{Medium}$, that will, eventually, transmit $m,ec$ to the $\symb{Receiver}$. 
%Strictly speaking, as variables are local to each  state, we have two distinct $ec$ variables in the \textit{Sender} pLTS, one in state \texttt{s1}, another one in state \texttt{s2}, the value of the variable is updated along the transitions entering each of the two states.

\begin{figure}[t]
   \includegraphics[width=.37\textwidth]{XFIG/SimpleProt-Schema.pdf}
   \includegraphics[width=.62\textwidth]{XFIG/SimpleProt2-Spec.pdf}
   \caption{pNet structure of the example and specification expressed as a pNet}
   \label{SimpleProt:Spec}

\end{figure}

\begin{figure}[t]
  \centerline{\includegraphics[width=.95\textwidth]{XFIG/SimpleProt2-pNet-tau}}
  \caption{Composed pNet with the Simple Protocol Implementation}  \label{SimpleProt:Impl}
\end{figure}

\subsection{Open Automata}
 Open automata (OA) are not composition structures but they are made of transitions that are dependent of the actions of the holes, and they can reason on a set of variables (potentially with only symbolic values). 
%\TODO{adopt a uniform notation for open transitions, almost each instance has a 
%different 
%notation! I suggest p,l,pr,po using \{\} for p,l,po as they are sets}
\begin{definition}[Open transitions]\label{def:OT}
	\label{def:OpenTransitions}
	An \emph{open transition} (OT) over a
	set $J$ of holes with sorts $\Sort_j^{j\in J}$, a set $V$ of variables, and a set of states $\mathcal{S}$ is 
	a structure of the form:	
	\begin{mathpar}
	\openrule
	{	\beta_j^{j\in J'}, \Pred, \Post}
	{s \OTarrow {\alpha}s'}
	\end{mathpar}
	Where $J'\subseteq J$, $s, s'\in\mathcal{S}$ and $\beta_j$
        is a transition of the hole $j$, with $\beta_j\in\Sort_j$. $\alpha$ is an action 
        label denoting the resulting action of this open transition.
%        We call \emph{source variables} of an OT the union of all variables
%        in the different terms
%        $\beta_j$ and $\alpha$, and the state variables of $s$.
        \Pred\ is a predicate over the  variables in $V$ and all variables
        in the different terms
       $\beta_j$ and $\alpha$. \Post\ is a set of 
	assignments that are effective after the open transition, they are
        represented as a substitution of the form $({x_k\gets e_k})^{k\in K}$ 
	where $\forall k.\, x_k\in V$, and $e_k$ are expressions
        over the variables $V$ and all variables
        in the different terms
       $\beta_j$ and $\alpha$. The assignments are always applied simultaneously because the variables in $V$ can be in both sides. Open transitions are identified
        modulo logical equivalence on their predicate. 
\end{definition}

An open automaton is  an automaton where each transition is an open transition.
\begin{definition}[Open automaton]
	\label{def:open-automaton}
	An \emph{open automaton} is a structure\\ $A =
	\mylangle J,\mathcal{S},s_0,V,\mathcal{T}\myrangle$ where:
	\begin{itemize}
		\item[$\bullet$]   $J$ is a  set of indices.
		\item[$\bullet$]   $\mathcal{S}$ is a set of states and $s_0$ an initial state
		  among $\mathcal{S}$.
%       \item[$\bullet$] for each state $s$, $vars(s)=\{v_s\}$ is a set of state variables, 
 \item[$\bullet$] $V$ is a set of variables of the automaton
		and each $v\in V$ may have an initial value $init(v)$.
		\item[$\bullet$] $\mathcal{T}$ is a set of open transitions and for each
		$t\in \mathcal{T}$ there exist  $J'$ with  $J'
		\subseteq J$, such that $t$ is an open transition over  $J'$
		and  $\mathcal{S}$.

We take in this article a semantics and logical understanding of these
automata. Open automata are closed by a simple form of refinement that
allows us to refine the predicate, or substitute any free variable by
an expression. More formally, let $\Pred$ be any predicate and $\Post$ any substitution such that $V\cap \dom(\Post)=\emptyset$. Then we have the following
implication:

	 \begin{mathpar}
    \openrule
         {
           \set{\beta}, \Pred\,',\Post\,'}
          {t \OTarrow {\alpha} t'}\in\mathcal{T}
\implies
    \openrule
         {
           \set{\beta}\subst{\Post}, \Pred\,'\subst{\Post}\land\Pred,\Post\shortotimes\Post\,'}
         {\raisemath{-2pt}{t \OTarrow {\alpha\subst{\Post}} {t'}}{}}
 \in\mathcal{T}
\end{mathpar}
\end{definition}

%\LUDO{I added this paragraph}
Because of the semantic interpretation of open automata, the set of open transition of an open automaton is infinite (for example because every free variable can be renamed). However an open automaton is characterized by a  subset of these open transition which is sufficient to generate, by substitution the other ones. In the following, we will abusively write that we define an ``open automaton'' when we provide only the set of open transitions that is sufficient to generate a proper open automaton by saturating each open transition by all possible substitutions and refinements.

 Then 
$\mathcal{R}$ is an FH-bisimulation if and only if for any  states
$s\in\mathcal{S}_1$ and $t\in\mathcal{S}_2$, $(s,t|\Pred_{s,t})\in\mathcal{R}$, we 
have
the following:
\end{minipage}
\hspace{2mm}
\begin{minipage}{0.30\linewidth}
	\includegraphics[width=\linewidth]{XFIG/Bisim.pdf}
\end{minipage}

\end{mathpar}
 there exist   open transitions $OT_x^{x\in X} \subseteq \mathcal{T}_2$:
 \begin{mathpar}
%    \left( fresh \ \set{\alpha_i}, \set{b_j}, v_x.\ \
    \openrule
         {
           \beta_{j x}^{j\in J_{x}}, \Pred_{OT_x},\Post_{OT_x}}
         {t \OTarrow {\alpha_x} t_x}
%         \right)
\end{mathpar}
 such that  $\forall x.\, J'=J_{x}$ and there exists $\Pred_{s',t_x}$ such that $(s',t_x|\Pred_{s',t_x})\in 
 \mathcal{R}$
 and  \\
 $\Pred_{s,t} \land \Pred_{OT}\implies$\\
%\hspace{1cm}
 $\displaystyle{\bigvee_{x\in X}
   \left( \forall j. \beta_j=\beta_{jx}  \land \Pred_{OT_x}
     \land \alpha\!=\!\alpha_x \land  
     \Pred_{s',t_x}\subst{\Post_{OT}\uplus\Post_{OT_x}}\right)}$
     %     \symb{Subst}(\Pred_{target_x}, \Post_{OT} o \Post_{OT_x})
     %     \right)$.
%\bigskip
%
% $\Pred_{s,t} \land \Pred_{OT}\implies\bigvee_{x\in X} \Pred_x$
%and\\
%%\hspace{1cm}
% $ \forall{x\in X}. \Pred_x \land \Pred_{s,t} \land \Pred_{OT} \Rightarrow
%   \left( \forall j. \beta_j=\beta_{jx}  \land \Pred_{OT_x}
%     \land \alpha\!=\!\alpha_x \land  
%     \Pred_{s',t_x}\subst{\Post_{OT}\uplus\Post_{OT_x}}\right)$
%     %     \symb{Subst}(\Pred_{target_x}, \Post_{OT} o \Post_{OT_x})
%     %     \right)$.

%     \TODO{Eric: j'ai ajoute $\exists$ sur les predicats $\Pred_{s',t_x}$, qui n'etaient pas definis...}

% \TODO{Eric: il reste des petits bugs genre $s^{2'}_x$ plutot que
%   $s^{2}_x$}

%Where $\symb{Subst}(\Pred,\Post)$ is the parallel substitution of all
%assigned variables.

% \TODO{do we want formulas for this?}
 \end{definition}
Classically, $\Pred_{s',t_x}\subst{\Post_{OT}\uplus\Post_{OT_x}}$
applies in parallel the  
substitutions $\Post_{OT}$ and $\Post_{OT_x}$ (parallelism is crucial
inside each $\Post$ set but not between  $\Post_{OT}$ and
$\Post_{OT_x}$ that are independent), applying the assignments of the involved rules.
We can prove that such a bisimulation is an equivalence relation.

%\TODO{have a look at the end of the sec}

@@

%\TODO{Eric: rewrite example with new Protocolkeep only OT1, and adapt comments}
\begin{example}[An open-transition]
  \label{OT:SimpleProt}
At the upper level, the $\symb{SimpleSystem}$ pNet of Figure \ref{SimpleProt:Impl} has 2 holes and $\symb{SimpleProtocol}$ as
a sub-pNet, itself containing 3 pLTSs. One of its possible open transitions
(synchronizing the hole $P$
with the $\symb{Sender}$ within the \emph{SimpleProtocol}) is:

 \smallskip\noindent
 \[  OT_1  = \openrule{
      \{\texttt{P}\mapsto \texttt{p-send(m)}\},  [\texttt{m=m'}],
        (\texttt{s\_msg}\gets \texttt{m})
                      }
    {\ostate{s_0,m_0,r_0} \OTarrow{\nounderline{\texttt{in(m')}}} \ostate{s_1,m_0,r_0}}
    \]

    \smallskip
    The global states here are triples, build as the product of states of the 3 pLTSs (remember the holes have no state). The assignment
    $Post$ uses the variable $\texttt{m}$ from the action of hole $\texttt{P}$ to set the
    value of the sender variable named $\texttt{s\_msg}$.

 %%  The \emph{SimpleProtocol} pNet of Fig. \ref{SimpleProt:Spec} has 3 controllers and no holes. One of its possible open-transition (the sender sending $(m)$ to the medium) is:

 %% \smallskip\noindent
 %% $  OT_2  = \openrule{
 %%   \emptyset; Pred=[send(SV\_m)=send(s_\_m);in(SV\_m)=in(?m)];  \\ Post=\{m_\_m:=?m\}
 %%                      }
 %%    {\ostate{s_1,m_0,r_0} \OTarrow{\nounderline{send(SV\_m)}} \ostate{s_2,m_1,r_0}}
 %%    $

 %%    That can be simplified to:

 %%    \ERIC{Apres application des substitutions et simplification du predicat:}

 %%     $  OT_2  = \openrule{
 %%      \emptyset; Pred=\True;  \\ Post=\{m_\_m:=s_\_m\}
 %%                      }
 %%    {\ostate{s_1,m_0,r_0} \OTarrow{\nounderline{send(s_\_m)}} \ostate{s_2,m_1,r_0}}
 %%    $

 %%    \smallskip
 %%    Note that the global states shows how the sender and the
 %%    first medium have evolved during this transition, the sender
 %%    moving from $s_1$ to $s_2$, and the first (forward) medium from
 %%    $m_0$ to $m_1$.
 %%    The predicate $Pred$ involves the state variables $s_\_m$ and
 %%    $s_\_sb$ of the sender, while the assignment $Post$ sets the
 %%    state variables of the medium.

    \smallskip

\end{example}

@@

%% \subsection{Post-processing: Elimination of intermediate variables}

%% While building the open transitions, many new variable names are
%% created, as fresh local variables of synchronisation vectors, and
%% fresh result variables of intermediate OTs. While these are required
%% for the semantic construction, they are not significant in the
%% resulting open automaton. In fact they carry information on the
%% internal structure of the pNet system, and this information should
%% \_not\_ be visible when looking at the system as a black box, and in
%% particular when comparing its behaviour with another system, e.g. by
%% bisimulation.

%% In the resulting predicate the only significant variables are :\\
%% - the input variables of pLTS transitions \TODO{heu, non, meme pas, si
%%   ce sont des comm internes!}\\
%% - the actions of holes\\
%% - the result action at toplevel.

%% So we transform the predicates of the resulting OTs, eliminating all
%% intermediate variables. \TODO{Explain why this should work !! Maybe
%%   need to be more precise on unification in the matching...}

%\TODO{Eric: full (???) result of the tool execution on the running example on the Simple Prot Spec: Done, on the Counetr version !}

%\begin{figure}[ht]
%   \centerline{\includegraphics[width=11cm]{ATG/SPSpecOpen}}
%   \caption{Specification OA}
%   \label{SimpleProtCounter:SpecOA}
%\end{figure}

\begin{figure}[ht]
   \centerline{\includegraphics[width=13cm]{XFIG/SPSpecOpen}}
   \caption{Specification Open Automaton}
   \label{SimpleProtCounter:SpecOA}
\end{figure}

\subsection*{Example}  Figure~\ref{SimpleProtCounter:SpecOA} shows the open automaton computed from the Specification pNet given in Figure \ref{SimpleProt:Spec}. 
For later references, we give to the transitions of this (strong)
Specification automaton as $SS_i$ while transitions of the
Implementation automaton will be labelled $SI_i$. In the figure we
have annotated each global state with the set of corresponding pLTS variables.

 \begin{figure}[ht]
  \centerline{\includegraphics[width=15cm]{XFIG/SPImplOpen}}
  \caption{Open Automaton of the Simple Protocol Implementation}  \label{SimpleProtCounter:ImplOA}
\end{figure}

>>>>

\documentclass[11pt]{article}
%\overfullrule=1mm
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{fullpage}
%\usepackage{geometry}                % See geometry.pdf to learn the layout options. There are lots.
%\geometry{a4paper}                   % ... or a4paper or a5paper or ...
%\geometry{landscape}                % Activate for for rotated page geometry
%\usepackage[parfill]{parskip}    % Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}
\usepackage{DotArrow}
\usepackage{hyperref}
\usepackage{listings}
\lstset{basicstyle=\small}

@@

%\pagestyle{plain}

@@

\title{ Open pNets
%\thanks{This work was partially funded by the Associated Team FM4CPS
%  between INRIA and ECNU, Shanghai}
}

@@

@@

For this property to be valid even if the substitution does not operate on all variables,
we define the composition operation as follows:

\[(x_{k}\gets e_{k})^{k\in K}\shortotimes (x'_{k'}\gets e'_{k'})^{k'\in K'} =  
(x_{k}\gets e_{k}\subst{(x'_{k'}\gets e'_{k'})^{k'\in K'}})^{k\in K} \cup (x'_{k'}\gets e'_{k'})^{k'\in K''}\]
where $K''=\{k'\in K'|x'_{k'}\not\in\{x_k\}^{k\in K}\}$

\subsection{Open Automata}
 Open automata (OA) are not composition structures but they are made of transitions that are dependent of the actions of the holes, and they can reason on a set of variables (potentially with only symbolic values). 
%\TODO{adopt a uniform notation for open transitions, almost each instance has a 
%different 
%notation! I suggest p,l,pr,po using \{\} for p,l,po as they are sets}
\begin{definition}[Open transitions]\label{def:OT}
	\label{def:OpenTransitions}
	An \emph{open transition} (OT) over a
	set $J$ of holes with sorts $\Sort_j^{j\in J}$, a set $V$ of variables, and a set of states $\mathcal{S}$ is 
	a structure of the form:	
	\begin{mathpar}
	\openrule
	{	\beta_j^{j\in J'}, \Pred, \phi}
	{s \OTarrow {\alpha}s'}
	\end{mathpar}
	Where $J'\subseteq J$, $s, s'\in\mathcal{S}$ and $\beta_j$
        is a transition of the hole $j$, with $\beta_j\in\Sort_j$. 
Let $W$ be the set of variables in $V$ and all variables
        in the different terms
       $\beta_j$ and $\alpha$.
$\alpha$ is an action 
        label denoting the resulting action of this open transition.
        \Pred\ is a predicate over the set of variables $W$. $\phi$, called \emph{effect} is a mapping from variables in $V$ to expressions over the set of variables $W$. $V$ is called the domain of the effect.
 Open transitions are identified
        modulo logical equivalence on their predicate. 
\end{definition}

Suppose $V=x_i^{i\in [1..n]}$ then $\phi$ has the form $(x_i\to e_i)^{i\in [1..n]}$.

Application of effect is as follows: $\phi(e)=(x_i\to e_i\subst{x_i\leftarrow e_i}$.

Composition of effects over the same domain is as follows:
Suppose $\phi=(x_i\to e_i)^{i\in [1..n]}$
and
$\psi=(x_i\to e'_i)^{i\in [1..n]}$.
Then $\phi(x'_i\to e'_i)^{i\in [1..m]} =(x'_i\to e'_i\subst{x_i\leftarrow e'_i}^{i\in [1..n]})^{i\in [1..m]}$. That applies to states or effects.

An open automaton is  an automaton where each transition is an open transition.
\begin{definition}[Open automaton]
	\label{def:open-automaton}
	An \emph{open automaton} is a structure\\ $A =
	\mylangle J,\mathcal{S},s_0,V,\mathcal{T}\myrangle$ where:
	\begin{itemize}
		\item[$\bullet$]   $J$ is a  set of indices.
		\item[$\bullet$]   $\mathcal{S}$ is a set of states and $s_0$ an initial state
		  among $\mathcal{S}$.
 \item[$\bullet$] $V$ is a set of variables of the automaton
		and each $v\in V$ may have an initial value $init(v)$.
		\item[$\bullet$] $\mathcal{T}$ is a set of open transitions and for each
		$t\in \mathcal{T}$ there exist  $J'$ with  $J'
		\subseteq J$, such that $t$ is an open transition over  $J'$
		and  $\mathcal{S}$.

We take in this article a semantics and logical understanding of these
automata. Open automata are closed by a simple form of refinement that
allows us to refine the predicate, or substitute any free variable by
an expression. More formally, let $\Pred$ be any predicate and $\psi$ any effect such that $V\cap \dom(\psi)=\emptyset$. Then we have the following
implication:

	 \begin{mathpar}
    \openrule
         {
           \set{\beta}, \Pred\,',\phi}
          {t \OTarrow {\alpha} t'}\in\mathcal{T}
\implies
    \openrule
         {
           \set{\psi(\beta)}, \psi(\Pred)\land\Pred,\phi\circ\psi}
         {\raisemath{-2pt}{t \OTarrow {\psi(\alpha)} {t'}}{}}
 \in\mathcal{T}
\end{mathpar}
\end{definition}

An operational interpretation of the open automaton and transition is naturally to build an automaton where states are couple (state identifier, variable valuation).
A variable valuation $\sigma$ maps variables to values. Similarly to effect, we use the notation $\sigma(e)$ to mean the evaluation of $e$ where variables have been instantiated by $\sigma$.
Consider  $A =
	\mylangle J,\mathcal{S},s_0,V,\mathcal{T}\myrangle$. Suppose $\dom(\sigma)=V$
We can build it for an open automaton without hole as follows (we obtain a concrete automaton $\mathcal{T'}$ from $\mathcal{T}$):

\begin{mathpar}
\inferrule
{\openrule
         {
           \emptyset, \Pred,\phi}
          {s \OTarrow {\alpha} s'}\in\mathcal{T}
\\
\dom(\psi)=\fv(\alpha)\setminus V
\\
(\psi\uplus\sigma)(\Pred)
}
{(s,\sigma)\xrightarrow{(\psi\uplus\sigma)(\alpha)}(s',\sigma\circ\phi\circ\psi)\in\mathcal{T'}}
\end{mathpar}

\TODO{We could add a set of automata in the state to take into account automata filling holes}

Composition: consider two rules obtained as above:
\begin{mathpar}
\inferrule
{(s,\sigma)\xrightarrow{(\psi\uplus\sigma)(\alpha)}(s',\sigma\circ\phi\circ\psi)\in\mathcal{T'}
\\
(s',\sigma\circ\phi\circ\psi)\xrightarrow{(\psi'\uplus\sigma')(\alpha')}(s'',\sigma\circ\phi\circ\psi\circ\phi'\circ\psi')\in\mathcal{T'}
}
{(s,\sigma)\xrightarrow{(\psi\uplus\sigma)(\alpha)~.~(\psi'\uplus\sigma\circ\phi\circ\psi)(\alpha')}(s'',\sigma\circ\phi\circ\psi\circ\phi'\circ\psi')\in\mathcal{T'}}
\end{mathpar}

Because of the semantic interpretation of open automata, the set of open transition of an open automaton is infinite (for example because every free variable can be renamed). However an open automaton is characterized by a  subset of these open transition which is sufficient to generate, by substitution the other ones. In the following, we will abusively write that we define an ``open automaton'' when we provide only the set of open transitions that is sufficient to generate a proper open automaton by saturating each open transition by all possible substitutions and refinements.

 Then 
$\mathcal{R}$ is an FH-bisimulation if and only if for any  states
$s\in\mathcal{S}_1$ and $t\in\mathcal{S}_2$, $(s,t|\Pred_{s,t})\in\mathcal{R}$, we 
have
the following:
\end{minipage}
\hspace{2mm}
\begin{minipage}{0.30\linewidth}
%	\includegraphics[width=\linewidth]{XFIG/Bisim.pdf}
\end{minipage}

\end{mathpar}
 there exist   open transitions $OT_x^{x\in X} \subseteq \mathcal{T}_2$:
 \begin{mathpar}
%    \left( fresh \ \set{\alpha_i}, \set{b_j}, v_x.\ \
    \openrule
         {
           \beta_{j x}^{j\in J_{x}}, \Pred_{OT_x},\Post_{OT_x}}
         {t \OTarrow {\alpha_x} t_x}
%         \right)
\end{mathpar}
 such that  $\forall x.\, J'=J_{x}$ and there exists $\Pred_{s',t_x}$ such that $(s',t_x|\Pred_{s',t_x})\in 
 \mathcal{R}$
 and  \\
 $\Pred_{s,t} \land \Pred_{OT}\implies$\\
%\hspace{1cm}
 $\displaystyle{\bigvee_{x\in X}
   \left( \forall j. \beta_j=\beta_{jx}  \land \Pred_{OT_x}
     \land \alpha\!=\!\alpha_x \land  
     \Pred_{s',t_x}\subst{\Post_{OT}\uplus\Post_{OT_x}}\right)}$

 \end{definition}
Classically, $\Pred_{s',t_x}\subst{\Post_{OT}\uplus\Post_{OT_x}}$
applies in parallel the  
substitutions $\Post_{OT}$ and $\Post_{OT_x}$ (parallelism is crucial
inside each $\Post$ set but not between  $\Post_{OT}$ and
$\Post_{OT_x}$ that are independent), applying the assignments of the involved rules.
We can prove that such a bisimulation is an equivalence relation.

In which $fv_{OT}$ denotes the set of free variables of all expressions in $OT$.

\newpage
\appendix

\section{Parameterised Networks (pNets)}
\label{section:pnets}

@@

@@

\section{pNet Composition Properties: composition of open transitions}
The semantics of open pNets allows us to prove two crucial properties relating pNet composition with pNet semantics: open transition of a composed pNet can be decomposed into open transitions of its composing sub-pNets, and conversely, from the open transitions of sub-pNets,  an open transition of the composed pNet can be built.

@@

\section{Bisimulation for open pNets -- a composable bisimulation theory}
\label{section:bisimulation-PN}
As  our symbolic operational semantics provides an open automaton, we can apply the notion of
	strong (symbolic) bisimulation on automata to open pNets:
\begin{definition}[FH-bisimulation for open pNets]\label{def:bisim-pnets}
Two pNets are FH-bisimilar if there exists a relation between their associated 
automata that is an FH-bisimulation and their initial states are in the relation (i.e. the predicate associated with the initial states is verifiable).
%\TODO{check}
\end{definition}
%\TODO{Predicate=True me pose un probleme... typiquement, on a besoin d'une ``condition initiale'', pour avoir des pNets data-oriented equivalents a leur version state-oriebted => remove the constraint}
